---
title: "Cytokine_data_analysis_Report"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: united
date: "2023-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Loading packages
```{r}

library(ggplot2)
library(patchwork)
library(gridExtra)
library(cowplot)


```

# Reading input files

```{r cars}

cytokine_whole_df <- read.table("Patient Data_Clustering_final.csv", sep=",", header=T)

head(cytokine_whole_df)
```

## Splitting data based on batches

Since the data was generated in different batches so the first thing would be to look for is batch effects.

```{r pressure}

# Adding "batch" to the first column of the cytokine_whole_df
cytokine_whole_df$BATCH <- paste("batch",cytokine_whole_df$BATCH,sep = "")

# Splitting the data based on different batches
splitted_df_list <- split(cytokine_whole_df, with(cytokine_whole_df, BATCH), drop = TRUE)

class(splitted_df_list)

# saving the data for batch_1 and batch_2 into cytokine_df_b1 and cytokine_df_b2
cytokine_df_b1 <- as.data.frame(splitted_df_list$batch1) 
cytokine_df_b2 <- as.data.frame(splitted_df_list$batch2)


```

# Removing batch effects
## function for making multiple plots

```{r plots}

create_density_plot <- function(df1,df2,column_name) {
   ggplot() +
    geom_density(data = df1, aes(x = .data[[column_name]], fill = "batch1"), alpha = 0.5) +
    geom_density(data = df2, aes(x = .data[[column_name]], fill = "batch2"), alpha = 0.5) +
    #labs(title = paste(column_name)) +
    #labs(fill = "Batch Name") 
    theme(legend.position="none") # Set the legend title

}

# Columns you want to compare
columns_to_compare <- c(colnames(cytokine_df_b1[,10:32]))

# Apply the function to each column
plt_before_scaling <- lapply(columns_to_compare, function(column) create_density_plot(cytokine_df_b1, cytokine_df_b2, column))
```

## normalizing/Scaling datasets to remove batch effects
```{r}

cytokine_df_b1[,10:32] <- as.data.frame(scale(cytokine_df_b1[,10:32]))
cytokine_df_b2[,10:32] <- as.data.frame(scale(cytokine_df_b2[,10:32]))


plt_after_scaling <- lapply(columns_to_compare, function(column) create_density_plot(cytokine_df_b1, cytokine_df_b2, column))
```

## The red color is batch 1 and blue color batch 2. The plots on left are shown before scaling/normalization and vice versa for the plots on the right

```{r}



n_plots <- length(plt_before_scaling)

# Create a new list of plots with added space between them
plots_with_space <- lapply(seq_len(n_plots), function(i) {
  grid.arrange(
    plt_before_scaling[[i]],
    plt_after_scaling[[i]],
    ncol = 2,
    widths = c(1, 1.5)  # Adjust the width of the second plot to add space
  )
})

# Arrange the new plots vertically
#grid.arrange(grobs = plots_with_space, ncol = 1)



```



# Combining the dataset after scaling and removing CCL7, CXCL1, CXCL11, GM.CSF, IL8, IL15, IL23, IL10 because of the all having same values

```{r}

combined_df <- rbind(cytokine_df_b1, cytokine_df_b2)

dim(combined_df)
class(combined_df)
head(combined_df)


# removing "CCL7", "CXCL1", "CXCL11", "GM.CSF", "IL8", "IL15", "IL23", "IL10"
combined_df <- combined_df[, !(names(combined_df) %in% c("CCL7", "CXCL1", "CXCL11", "GM.CSF", "IL8", "IL15", "IL23", "IL10"))]

dim(combined_df)
class(combined_df)
head(combined_df)


# changing column names for treatment and KRAS mutation
colnames(combined_df)[8:9] <- c("treatment","KRAS_mutation")


# setting clinical.ID in the dataframe as the row names or the sample name for the dataframe
row.names(combined_df) <- combined_df$Clinical.ID
combined_df <- combined_df[,-2]

colnames(combined_df)
rownames(combined_df)
```



# kmeans clustering and UMAP Visualization 


## Loading required packages
```{r}
library(FactoMineR)
library(stats)
library(umap)
library(factoextra)
library(RColorBrewer)
library(purrr)
```

## Performing PCA
```{r}
# Process the dataset
columns_for_pca <- combined_df[,9:23]

dim(columns_for_pca)

columns_for_pca <- na.omit(columns_for_pca)

dim(columns_for_pca)

# Step 1: PCA
pca_result <- PCA(columns_for_pca, scale.unit = FALSE, graph = FALSE, ncp = 5)

# Scree_plot
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 50))

```

## selecting the appropriate number of clusters 
```{r}
# Determine the number of clusters using fb_nviz
fviz_nbclust(pca_result$ind$coord, kmeans, method = "wss")

fviz_contrib(pca_result,
             choice = "var",
             axes = 1,
             top = 10, color = 'darkorange3', barfill  = 'blue4',fill ='blue4')
```

## top 10 cytokines contributing to PC1 and PC2
```{r}
fviz_contrib(pca_result,
             choice = "var",
             axes = 2,
             top = 10, color = 'darkorange3', barfill  = 'blue4',fill ='blue4')
# Create data frames for plotting
# Add PCA, kmeans, and UMAP results to combined_df

non_missing_rows <- rownames(columns_for_pca)

# Extract corresponding rows from the original combined_df
combined_df_no_missing <- combined_df[non_missing_rows, ]

combined_df_no_missing$PCA1 <- pca_result$ind$coord[, 1]
combined_df_no_missing$PCA2 <- pca_result$ind$coord[, 2]

```



## Function to perform k-means clustering and create UMAP plot

```{r}

perform_kmeans_umap <- function(df, pca_result, n_pcs) {
  
  k_values <- c(2,3, 4, 5)
  
  # Generate distinct colors
  num_colors <- max(k_values)
  colours <- brewer.pal(num_colors, "Set1")
  
  
     
  # Perform UMAP
  umap_result <- umap(pca_result$ind$coord[, 1:n_pcs])

  # Save UMAP coordinates in the dataframe
  df$UMAP1 <- umap_result$layout[, 1]
  df$UMAP2 <- umap_result$layout[, 2]
  
  # Perform k-means clustering for different values of k using purrr::map
  umap_plots <- map(k_values, ~{
    kmeans_result <- kmeans(pca_result$ind$coord[, 1:n_pcs], centers = .x)
    df[[paste0("Cluster_k", .x)]] <- kmeans_result$cluster
    
    ggplot(df, aes(x = UMAP1, y = UMAP2, shape = factor(treatment))) +
      geom_point(aes(color = factor(df[[paste0("Cluster_k", .x)]])), size = 3) +
      scale_color_manual(values = colours) +
      scale_shape_manual(values = c(0, 17)) +
      labs(title = paste("UMAP Visualization with K-means Clustering (k=", .x, ")"))
  })

  names(umap_plots) <- paste0("UMAP_k", k_values)

  return(umap_plots)
}


# Example usage: specify the number of principal components and k values
n_pcs <- 4


# Call the function
umap_results <- perform_kmeans_umap(combined_df_no_missing, pca_result, n_pcs)

```




## kmeans clustering UMAP with 4 PCs

```{r}

# Example usage: specify the number of principal components and k values
n_pcs <- 4


# Call the function
umap_results <- perform_kmeans_umap(combined_df_no_missing, pca_result, n_pcs)

```

## UMAPS {.tabset}

### UMAP with 2 clusters
```{r}

umap_results[1]
write.table(data.frame(umap_results[["UMAP_k2"]]$data),"cytokine_info_file_whole_data_PC4_k2.txt", sep="\t",quote = F)

```


### UMAP with 3 clusters
```{r}

umap_results[2]
write.table(data.frame(umap_results[["UMAP_k3"]]$data),"cytokine_info_file_whole_data_PC4_k3.txt", sep="\t",quote = F)
```


### UMAP with 4 clusters
```{r}

umap_results[3]

```

### UMAP with 5 clusters
```{r}

umap_results[4]

```

{-}

## kmeans clustering UMAP with 2 PCs

```{r}

# Example usage: specify the number of principal components 
n_pcs <- 2

# Call the function
umap_results <- perform_kmeans_umap(combined_df_no_missing, pca_result, n_pcs)
```







## UMAPS {.tabset}

### UMAP with 2 clusters
```{r}

umap_results[1]

write.table(data.frame(umap_results[["UMAP_k2"]]$data),"cytokine_info_file_whole_data_PC2_k2.txt", sep="\t",quote = F)

table(umap_results$UMAP_k2$data$Cluster_k2)
```


### UMAP with 3 clusters
```{r}

umap_results[2]
write.table(data.frame(umap_results[["UMAP_k3"]]$data),"cytokine_info_file_whole_data_PC2_k3.txt", sep="\t",quote = F)

```


### UMAP with 4 clusters
```{r}

umap_results[3]

```

### UMAP with 5 clusters
```{r}

umap_results[4]

```

{-}


# UMAP plots with 2 PCs (and two clusters) for various conditions such as sex, Kras mutation and 
```{r}


# making a dataframe for UMAP with 2 PCs and 2 clusters
df_k2_pC2 <-  as.data.frame(umap_results[["UMAP_k2"]]$data)

# writing this dataframe into file
write.table(df_k2_pC2, "cytokine_normalized_data.txt", sep="\t", quote=F)

# making plots for response
ggplot(df_k2_pC2, aes(x = UMAP1, y = UMAP2, color = factor(Cluster_k2), shape = Response)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("blue", "red")) +  # Adjust colors as needed 
  labs(title = "UMAP Plot with Clustering Information")


# making plots for Sex
ggplot(df_k2_pC2, aes(x = UMAP1, y = UMAP2, color = factor(Cluster_k2), shape = Sex)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("blue", "red")) +  # Adjust colors as needed 
  labs(title = "UMAP Plot with Clustering Information")


# making plots for Kras mutation
ggplot(df_k2_pC2, aes(x = UMAP1, y = UMAP2, color = factor(Cluster_k2), shape = KRAS_mutation)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("blue", "red")) +  # Adjust colors as needed 
  labs(title = "UMAP Plot with Clustering Information")


```


# Performing the separate analysis for monotherapy and combo therapy
## Splitting the dataset based on the two treatment options
```{r}

# the number of monotherapy and combo_therapy in the whole dataframe
table(combined_df_no_missing$treatment)

monotherapy_df <- combined_df_no_missing[combined_df_no_missing$treatment == "Mono", ]
combo_therapy_df <- combined_df_no_missing[combined_df_no_missing$treatment == "Combo", ]

dim(monotherapy_df)





```

## Analysis for splitted datasets
```{r}


perform_pca_kmeans_umap <- function(df, n_pcs, k_values) {
  # Process the dataset
  columns_for_pca <- df[, 9:23]
  columns_for_pca <- na.omit(columns_for_pca)

  # Step 1: PCA
  pca_result <- PCA(columns_for_pca, scale.unit = FALSE, graph = FALSE, ncp = n_pcs)

  # Selecting the appropriate number of clusters
  fviz_nbclust(pca_result$ind$coord, kmeans, method = "wss")
  
  
  print(fviz_contrib(pca_result,
             choice = "var",
             axes = 1,
             top = 10, color = 'darkorange3', barfill  = 'blue4',fill ='blue4'))
  
  
  print(fviz_contrib(pca_result,
             choice = "var",
             axes = 2,
             top = 10, color = 'darkorange3', barfill  = 'blue4',fill ='blue4'))
  # Create data frames for plotting
  non_missing_rows <- rownames(columns_for_pca)
  df_no_missing <- df[non_missing_rows, ]

  # Extract corresponding rows from the original dataframe
  df_no_missing$PCA1 <- pca_result$ind$coord[, 1]
  df_no_missing$PCA2 <- pca_result$ind$coord[, 2]

  # Perform k-means clustering and create UMAP plot
  umap_results <- perform_kmeans_umap(df_no_missing, pca_result, n_pcs, k_values)

  return(umap_results)
}

# Function to perform k-means clustering and create UMAP plot
perform_kmeans_umap <- function(df, pca_result, n_pcs, k_values) {
  # Generate distinct colors
  num_colors <- max(k_values)
  colours <- brewer.pal(num_colors, "Set1")

  # Perform UMAP
  umap_result <- umap(pca_result$ind$coord[, 1:n_pcs])

  # Save UMAP coordinates in the dataframe
  df$UMAP1 <- umap_result$layout[, 1]
  df$UMAP2 <- umap_result$layout[, 2]

  # Perform k-means clustering for different values of k using purrr::map
  umap_plots <- map(k_values, ~{
    kmeans_result <- kmeans(pca_result$ind$coord[, 1:n_pcs], centers = .x)
    df[[paste0("Cluster_k", .x)]] <- kmeans_result$cluster

    ggplot(df, aes(x = UMAP1, y = UMAP2, shape = factor(Response))) +
      geom_point(aes(color = factor(df[[paste0("Cluster_k", .x)]])), size = 3) +
      scale_color_manual(values = colours) +
      scale_shape_manual(values = c(0, 17,2)) +
      labs(title = paste("UMAP Visualization with K-means Clustering (k=", .x, ")"))
  })

  names(umap_plots) <- paste0("UMAP_k", k_values)

  return(umap_plots)
}

# Example usage
n_pcs <- 2
k_values <- c(2, 3, 4)



```

### monotherapy analysis results
### Contributions of top 10 cytokines to PC1 and PC2
```{r}
# Perform PCA, k-means clustering, and create UMAP plots for monotherapy

umap_results_monotherapy <- perform_pca_kmeans_umap(monotherapy_df, n_pcs, k_values)

```

### UMAPS {.tabset}

#### with two clusters
```{r}

umap_results_monotherapy[1]
write.table(data.frame(umap_results_monotherapy[["UMAP_k2"]]$data),"cytokine_info_file_monotherapy_PC2_k2.txt", sep="\t",quote = F)

```

#### with three clusters
```{r}
umap_results_monotherapy[2]
write.table(data.frame(umap_results_monotherapy[["UMAP_k3"]]$data),"cytokine_info_file_monotherapy_PC2_k3.txt", sep="\t",quote = F)
```

#### with four clusters
```{r}
umap_results_monotherapy[3]
```
{-}

### combo_therapy analysis results
### Contributions of top 10 cytokines to PC1 and PC2 
```{r}
# Perform PCA, k-means clustering, and create UMAP plots for combo therapy
umap_results_combo_therapy <- perform_pca_kmeans_umap(combo_therapy_df, n_pcs, k_values)
```

### UMAPS {.tabset}

#### with two clusters
```{r}

umap_results_combo_therapy[1]
write.table(data.frame(umap_results_combo_therapy[["UMAP_k2"]]$data),"cytokine_info_file_combo_therapy_PC2_k2.txt", sep="\t",quote = F)
```

#### with three clusters
```{r}
umap_results_combo_therapy[2]
write.table(data.frame(umap_results_combo_therapy[["UMAP_k3"]]$data),"cytokine_info_file_combotherapy_PC2_k3.txt", sep="\t",quote = F)

```

#### with four clusters
```{r}
umap_results_combo_therapy[3]
```

# Performing the separate analysis for males and females
## Splitting the dataset based on the two treatment options
```{r}



male_df <- combined_df_no_missing[combined_df_no_missing$Sex == "M", ]
female_df <- combined_df_no_missing[combined_df_no_missing$Sex == "F", ]


# printing the dimensions of the male dataframe
dim(male_df)


# printing the dimensions of the female dataframe
dim(female_df)

```

### perform analysis for males
## top 10 cytokines contributing to PC1 and PC2
```{r}
 # changing the number of clusters to 4
n_pcs <- 4
umap_results_male <- perform_pca_kmeans_umap(male_df, n_pcs, k_values)

```


### UMAPS {.tabset}

#### with two clusters
```{r}
umap_results_male[1]
write.table(data.frame(umap_results_male[["UMAP_k2"]]$data),"cytokine_info_file_male_PC4_k2.txt", sep="\t",quote = F)
```

#### with three clusters
```{r}
umap_results_male[2]
write.table(data.frame(umap_results_male[["UMAP_k3"]]$data),"cytokine_info_file_male_PC4_k3.txt", sep="\t",quote = F)

```

#### with four clusters
```{r}
umap_results_male[3]
```


### perform analysis for females
## top 10 cytokines contributing to PC1 and PC2
```{r}

umap_results_female <- perform_pca_kmeans_umap(female_df, n_pcs, k_values)

```



### UMAPS {.tabset}

#### with two clusters
```{r}
umap_results_female[1]
write.table(data.frame(umap_results_female[["UMAP_k2"]]$data),"cytokine_info_file_female_PC4_k2.txt", sep="\t",quote = F)
```

#### with three clusters
```{r}
umap_results_female[2]
write.table(data.frame(umap_results_female[["UMAP_k3"]]$data),"cytokine_info_file_female_PC4_k3.txt", sep="\t",quote = F)

```

#### with four clusters
```{r}
umap_results_female[3]
```
