---
title: "Geo_MX Data Analysis"
output: html_document
date: "2024-02-28"
Author: "Ammar Sabir Cheema"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading input files
```{r cars}
library(rio) 
path <- "/mnt/acri4_1/acheema/geo_mx_analysis/Initial_Dataset_Maha.xlsx"
spatial_data <- import_list(path)  


sampleAnnoFile <- spatial_data$SegmentProperties
head(sampleAnnoFile)


write.table(sampleAnnoFile,
             "metadata.txt", 
              row.names = T, 
              col.names = T, 
              quote = F, 
              sep = '\t')

featureAnnoFile <- spatial_data$BioProbeCountMatrix[, c("TargetName" ,"HUGOSymbol" ,"TargetGroup", "AnalyteType"   ,   "CodeClass") ]
head(featureAnnoFile)


write.table(featureAnnoFile,
            "genemeta.txt", 
              row.names = T, 
              col.names = T, 
              quote = F, 
              sep = '\t')

countFile <- spatial_data$BioProbeCountMatrix[, c(3 ,13:108) ]
head(countFile)


write.table(  countFile,
              "count.txt", 
              row.names = T, 
              col.names = T, 
              quote = F, 
              sep = '\t')
```


# Analyzing data, loading packages, making spe (summarized experiment object)

```{r pressure}
library(standR)


spe <- readGeoMx(countFile, sampleAnnoFile, featureAnnoFile)
```


# Loading spe to look into its spe
```{r}


spe
```

# Exploratory Data Anaysis for the spe object using spatialexperiment package
```{r}

library(SpatialExperiment)

assayNames(spe)


assay(spe, "counts")[1:5,1:5]


assay(spe, "logcounts")[1:5,1:5]
```


# printing the sample metadata and the gene metadata information for the spe object
```{r}

head(colData(spe))

head(rowData(spe))
```

# negavtive probes information
```{r}


metadata(spe)$NegProbes[,1:5]


spe@colData$SlideName = spe@colData$ROILabel

```


# Import from DGEList object
```{r}

dge <- edgeR::SE2DGEList(spe)

spe2 <- readGeoMxFromDGE(dge)

spe2

```

# QC Flags
```{r}

colData(spe)$QCFlags

```

# Quality Controls
```{r}

library(ggplot2)
library(ggalluvial)

plotSampleInfo(spe, column2plot = c("SlideName","SegmentLabel"))


```


# Gene level QC
```{r}
dim(spe)

spe <- addPerROIQC(spe, rm_genes = TRUE)


dim(spe)


metadata(spe) |> names()

```

# plot gene qc
```{r}
plotGeneQC(spe, ordannots = "regions", col = regions, point_size = 2)

```


# ROI level QC
```{r}

plotROIQC(spe, x_threshold = 150, color = SlideName)


qc <- colData(spe)$AOINucleiCount > 150

table(qc)
#FALSE  TRUE 
#   15    81
spe <- spe[, qc]
dim(spe)
```


# threshold based on library size
```{r}

plotROIQC(spe,  x_axis = "AOISurfaceArea", x_lab = "AreaSize", y_axis = "lib_size", y_lab = "Library size", col = SlideName)

```

# Relative log expression distribution
```{r}

plotRLExpr(spe)



plotRLExpr(spe, ordannots = "SlideName", assay = 2, color = SlideName)






plotRLExpr(spe, ordannots = "SegmentLabel", assay = 2, color = SegmentLabel)
```


# Dimenionsionality Reduction
```{r}


drawPCA(spe, assay = 2, color = SegmentLabel)
```


# calculating the pca using scater to have consistent results
```{r}


set.seed(100)

spe <- scater::runPCA(spe)

pca_results <- reducedDim(spe, "PCA")

drawPCA(spe, precomputed = pca_results, col = SegmentLabel)
```

# Again getting only one slide name for SlideName
```{r}

drawPCA(spe, precomputed = pca_results, col = SlideName)

```


# making scree plot
```{r}

plotScreePCA(spe, precomputed = pca_results)
```


# making plot in pairs
```{r}


plotPairPCA(spe, col = SegmentLabel, precomputed = pca_results, n_dimension = 4)

```


# PCA
```{r}

plotPairPCA(spe, col = SlideName, precomputed = pca_results, n_dimension = 4)



plotPCAbiplot(spe, n_loadings = 10, precomputed = pca_results, col = SegmentLabel)
```


# making MDS plot
```{r}
standR::plotMDS(spe, assay = 2, color = SegmentLabel)
```


# umap
```{r}

set.seed(100)

spe <- scater::runUMAP(spe, dimred = "PCA")

plotDR(spe, dimred = "UMAP", col = SegmentLabel)

```


# Normalization
```{r}
spe_tmm <- geomxNorm(spe, method = "TMM")
```


```{r}

plotRLExpr(spe_tmm, assay = 2, color = SlideName) + ggtitle("TMM")


set.seed(100)

spe_tmm <- scater::runPCA(spe_tmm)

pca_results_tmm <- reducedDim(spe_tmm, "PCA")

plotPairPCA(spe_tmm, precomputed = pca_results_tmm, color = SegmentLabel)
```

# this part is not working, it asks to put rmNegProbe=FALSE, but when I put it FALSE it gives error that rownames are duplicated.  
```{r}

# library(SpatialDecon)
# library(SummarizedExperiment)
# 
# spe <- readGeoMx(countFile, sampleAnnoFile, featureAnnoFile, rmNegProbe=TRUE)
# 
# spe <- addPerROIQC(spe, rm_genes = TRUE)
# 
# qc <- colData(spe)$AOINucleiCount > 150
# 
# spe <- spe[, qc]
# 
# spe_tmm <- geomxNorm(spe, method = "TMM")
# spd <- prepareSpatialDecon(spe_tmm)
# 
# 
#  names(spd)

```


# unsupervised clustering 
```{r}
library(factoextra)

table(duplicated(countFile$TargetName))
							 countFile <- countFile[!duplicated(countFile$TargetName, fromLast = TRUE), ]
							 countFile$TargetName <- NULL
							 
							 
							 # adding rownames to the dataframe
							 rownames(countFile) <- countFile$TargetName
							 
							 
							 # take the transpose of the dataframe
							 countFile <- as.data.frame(t(countFile))
							
							# extracting the tissue names from the rownames
							countFile$tissue_names <- sapply(rownames(countFile), function(x) tail(strsplit(x, "\\|")[[1]], 1))
							  
						    # split the dataframe based on tissue names
							split_dfs <- split(countFile, countFile$tissue_names)
							
    					
							
							get_top_variable_genes <- function(df) {
                           variances <- apply(df, 2, var,na.rm=TRUE)  # Calculate variance for each gene across samples
                           topGenes <- names(sort(variances, decreasing = TRUE))[1:1000]  # Get top 1000 genes by variance
                           return(topGenes)
    					   }
    					   
    					   
    					   
							 # getting the top variable genes for all the tissue types and saving in the list
						   top_variable_genes_list <- lapply(split_dfs, get_top_variable_genes)
						   
						   # printing the head of the list of gene in top_variable_genes_list
						   lapply(top_variable_genes_list, head, n = 50)
							
						   
                           get_top_variable_genes_df <- function(df, top_genes) {
                                  return(t(df[,top_genes ]))
                           }
                           
                           
                           top_variable_genes_df_list <- lapply(seq_along(split_dfs), function(i) {
                           get_top_variable_genes_df(split_dfs[[i]], top_variable_genes_list[[i]])
                           })

                           
                           
                           
                           
                           

							  perform_kmeans <- function(df,max_k = 5) {
	  # Assuming gene expression data starts from the second column
	  gene_expr <- df  # Exclude the first column (assuming it's non-gene data)
	  
	  
	  print(fviz_nbclust(gene_expr, kmeans, method = "silhouette", k.max = max_k))  
	  
	  # Perform k-means clustering
	  res.km <- kmeans(scale(gene_expr), 3, nstart = 25)  # You can adjust the number of centers as needed
	  
	  # K-means clusters showing the group of each individuals
	  res.km$cluster
	  
	  print(fviz_cluster(res.km, data = gene_expr,
				 palette = c("#2E9FDF", "#00AFBB", "#E7B800"), 
				 geom = "point",
				 ellipse.type = "convex", 
				 ggtheme = theme_bw()
				 ))
	  
	  return(res.km)
	}

	# List to store k-means clustering results for each dataframe
	kmeans_results_list <- lapply(top_variable_genes_df_list, perform_kmeans)
	
	


```
